---
title: "Exploratory Data Analysis — Films"
author: "Jakub Groblicki"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: hide
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5
)
set.seed(42)
```

```{css, echo=FALSE}
body { font-size: 15px; line-height: 1.8; }
h1 { border-bottom: 2px solid #2c3e50; padding-bottom: 6px; margin-top: 2rem; }
h2 { color: #2c3e50; margin-top: 1.8rem; }
h3 { color: #6d597a; margin-top: 1.4rem; }
blockquote {
  border-left: 4px solid #6d597a;
  background: #f5f0f8;
  padding: 12px 18px;
  margin: 16px 0;
  border-radius: 4px;
  font-size: 14.5px;
}
.result-box {
  background: #eaf4ea;
  border-left: 4px solid #2a9d8f;
  padding: 12px 18px;
  border-radius: 4px;
  margin: 16px 0;
  font-size: 14.5px;
}
.note-box {
  background: #fef9ec;
  border-left: 4px solid #e9c46a;
  padding: 12px 18px;
  border-radius: 4px;
  margin: 16px 0;
  font-size: 14.5px;
}
p { margin-bottom: 0.9rem; }
```

```{r packages, include=FALSE}
library(tidyverse)
library(psych)
library(gridExtra)
library(mice)
library(corrplot)
library(nortest)
library(scales)
library(dunn.test)
```

```{r palette, include=FALSE}
col_profit <- "#2a9d8f"
col_loss   <- "#e76f51"
col_nom    <- "#6d597a"
col_none   <- "#a8dadc"
col_gold   <- "#e9c46a"
```

```{r helper-awards, include=FALSE}
assign_award_group <- function(df,
                               col_awards = "Oscar and Golden Globes awards",
                               col_noms   = "Oscar and Golden Globes nominations",
                               labels     = c("No nominations",
                                              "Nominated, no award",
                                              "At least 1 award")) {
  df %>%
    mutate(
      award_group = factor(
        ifelse(.data[[col_awards]] > 0, labels[3],
               ifelse(.data[[col_noms]] > 0, labels[2], labels[1])),
        levels = labels
      )
    )
}

effect_label <- function(rho) {
  ifelse(abs(rho) < 0.10, "negligible",
  ifelse(abs(rho) < 0.30, "small",
  ifelse(abs(rho) < 0.50, "medium", "large")))
}
```

---

# Introduction

This report presents an Exploratory Data Analysis (EDA) of a dataset of
feature films. The analysis examines financial outcomes (budget, box office
revenue, and earnings), IMDb audience ratings, and selected descriptive
attributes of productions including genre, release year, director, and cast.

The analysis is structured around three research questions:

1. **What factors are associated with a film posting a negative financial result?**
2. **What influences a film's IMDb rating?**
3. **Do industry-recognised films differ systematically from the rest?**

All findings are exploratory — they describe patterns, distributions and
associations observed in the data, and do not imply causal relationships.

---

# Dataset

## Description

The dataset contains information on feature films drawn from several decades
of theatrical releases. Each row corresponds to one film and includes
financial figures, IMDb scores, and descriptive attributes.

| Variable | Type | Description |
|---|---|---|
| `Movie` | Text | Film title |
| `Director` | Text | Director's name |
| `Running time` | Numeric | Duration (minutes) |
| `Actor 1` | Text | Lead actor |
| `Actor 2` | Text | Supporting actor |
| `Actor 3` | Text | Third-billed actor |
| `Genre` | Categorical | Genre (e.g. Action, Comedy, Drama) |
| `Budget` | Numeric | Production budget (USD) |
| `Box Office` | Numeric | Total theatrical revenue (USD) |
| `Actors Box Office %` | Numeric | % of the cast's prior films where box office ≥ 2× budget |
| `Director Box Office %` | Numeric | % of the director's prior films where box office ≥ 2× budget |
| `Earnings` | Numeric | Profit = Box Office − Budget (USD) |
| `Oscar and Golden Globes nominations` | Numeric | Total Oscar and Golden Globe nominations received |
| `Oscar and Golden Globes awards` | Numeric | Total Oscars and Golden Globes won |
| `Release year` | Numeric | Year of theatrical release |
| `IMDb score` | Numeric | Mean IMDb audience rating (scale 1–10) |

## Loading the Data

```{r load}
movies <- readr::read_csv(
  "movies_data.csv",
  locale = readr::locale(encoding = "windows-1252")
)
```

A preview of the first and last rows confirms the general structure and
format of the data:

```{r preview}
head(movies, 5)
tail(movies, 5)
```

```{r structure}
str(movies)
```

The dataset contains `r nrow(movies)` rows and `r ncol(movies)` columns.
Financial variables are stored as integers (USD values with no decimal
places), which is consistent with expectations.

---

## Data Cleaning

### Missing values

```{r missing-check}
anyNA(movies)
colSums(is.na(movies))
```

```{r missing-plot, fig.height=4}
md.pattern(movies)
```

Missing values are confined to two columns. `Actor 3` has
`r sum(is.na(movies$"Actor 3"))` missing entries — a third-billed actor is
simply absent for many productions, so these are replaced with `"Unknown"`.
`Oscar and Golden Globes awards` has
`r sum(is.na(movies$"Oscar and Golden Globes awards"))` missing entries.
Before imputing, we verify that all affected films also have zero
nominations — which would confirm the missing value means *no award won*
rather than a data collection failure:

```{r missing-impute}
na_awards_idx <- is.na(movies$`Oscar and Golden Globes awards`)
cat("Nomination counts for films with missing award values:\n")
table(movies$`Oscar and Golden Globes nominations`[na_awards_idx])

movies$`Oscar and Golden Globes awards`[na_awards_idx] <- 0
movies$`Actor 3`[is.na(movies$`Actor 3`)] <- "Unknown"
```

All films with a missing award count also have zero nominations, confirming
that the missing values represent absent awards rather than absent data.
Both columns are imputed accordingly.

### Duplicate rows

```{r duplicates}
cat("Duplicate rows found:", sum(duplicated(movies)), "\n")
movies <- movies[!duplicated(movies), ]
cat("Rows retained after deduplication:", nrow(movies), "\n")
```

---

## Data Preparation

### Type conversion

`Genre` is converted to a factor to ensure correct grouping behaviour and
ordered axis labels in plots:

```{r genre-factor}
movies$Genre <- as.factor(movies$Genre)
cat("Genres present:", paste(levels(movies$Genre), collapse = ", "), "\n")
```

The dataset contains `r nlevels(movies$Genre)` distinct genres.

### Verifying `Earnings`

We confirm that `Earnings` equals Box Office − Budget for every row, and
that it does not incorporate other revenue streams such as rights sales or
merchandise:

```{r earnings-check}
all_zero <- all((movies$Earnings - (movies$`Box Office` - movies$Budget)) == 0)
cat("Earnings equals Box Office minus Budget for all rows:", all_zero, "\n")
```

The check confirms a perfect identity. `Earnings` reflects theatrical
revenues only — a point worth bearing in mind throughout the financial
analysis, since real profitability would also depend on home release,
streaming rights, and ancillary markets.

### Adding the ROI variable

Raw earnings are scale-dependent: a \$300M profit means something very
different on a \$30M budget than on a \$250M budget. We add a **Return on
Investment** ratio (`roi = Box Office / Budget`) to capture relative
financial performance alongside absolute earnings:

```{r roi}
movies <- movies %>% mutate(roi = `Box Office` / Budget)
summary(movies$roi)
```

A film with `roi > 1` recovered its production costs from theatrical
revenues alone. An `roi > 2` is a common industry benchmark for a
commercially successful release.

### Correcting `Actors Box Office %`

`Actors Box Office %` is a percentage (range 0–100) representing the
proportion of an actor's prior films where box office reached at least 2×
the production budget. Any value above 100 is logically impossible and
indicates a data entry error:

```{r actors-fix}
cat("Records with Actors Box Office % > 100:",
    sum(movies$`Actors Box Office %` > 100, na.rm = TRUE), "\n")

movies$`Actors Box Office %`[movies$`Actors Box Office %` > 100] <- NA

cat("Records with Actors Box Office % > 100 after correction:",
    sum(movies$`Actors Box Office %` > 100, na.rm = TRUE), "\n")
```

Both erroneous records are replaced with `NA`.

### Pre-computing director performance groups

To avoid repeating the same `cut()` call across multiple sections, the
director performance quartile grouping is computed once here and reused
throughout:

```{r director-groups}
movies <- movies %>%
  mutate(
    director_group = cut(
      `Director Box Office %`,
      breaks = quantile(`Director Box Office %`,
                        probs = c(0, .25, .5, .75, 1), na.rm = TRUE),
      labels = c("Low (Q1)", "Mid-low (Q2)", "Mid-high (Q3)", "High (Q4)"),
      include.lowest = TRUE
    )
  )
```

---

## Descriptive Statistics

```{r descriptive-stats}
num_cols <- c("Running time", "Budget", "Box Office",
              "Actors Box Office %", "Director Box Office %",
              "Earnings", "roi",
              "Oscar and Golden Globes nominations",
              "Oscar and Golden Globes awards",
              "Release year", "IMDb score")

summary(movies[, num_cols])
```

The `psych` package extends the summary with shape measures — skewness and
kurtosis:

```{r describe}
describe(movies[, num_cols]) %>%
  select(n, mean, sd, median, min, max, skew, kurtosis) %>%
  round(2)
```

<div class="note-box">
**Kurtosis convention.** `psych::describe()` reports *excess kurtosis*
(standard Pearson kurtosis minus 3). A normal distribution therefore has
excess kurtosis = 0. Positive values denote heavier tails than normal
(leptokurtic); negative values denote lighter tails (platykurtic).
</div>

The most striking feature of the descriptive table is the strong
right-skewness of all financial variables. `Budget` and `Box Office` exhibit
very long right tails — a pattern typical of entertainment industry data,
where revenue is highly concentrated around a few major blockbusters.
`Earnings` is similarly asymmetric, centred around small positive values
with a pronounced right tail from major hits and a visible left tail from
loss-making productions. This distributional shape has a direct
methodological implication: **robust location measures (median, IQR) and
non-parametric inferential tests are the appropriate tools for all financial
variables** throughout this analysis.

### Distributions of financial variables

The three histograms below illustrate the key financial variables. Budget
and Box Office use a logarithmic x-axis — without it, a handful of extreme
outliers would compress the entire distribution into an unreadable spike.
Earnings cannot be log-transformed because it takes negative values, so the
raw scale is used instead, with bars coloured by profit/loss status:

```{r financial-distributions, fig.height=4}
p1 <- ggplot(movies, aes(x = Budget)) +
  geom_histogram(fill = col_nom, colour = "white", bins = 40) +
  scale_x_log10(labels = label_dollar(scale = 1e-6, suffix = " M")) +
  labs(title = "Budget", x = "Budget [USD, log scale]",
       y = "Number of films") +
  theme_minimal(base_size = 11)

p2 <- ggplot(movies, aes(x = `Box Office`)) +
  geom_histogram(fill = col_profit, colour = "white", bins = 40) +
  scale_x_log10(labels = label_dollar(scale = 1e-6, suffix = " M")) +
  labs(title = "Box Office", x = "Box office [USD, log scale]",
       y = "Number of films") +
  theme_minimal(base_size = 11)

p3 <- ggplot(movies, aes(x = Earnings)) +
  geom_density(fill = col_nom, alpha = 0.6, colour = col_nom) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey30", linewidth = 0.7) +
  scale_x_continuous(
    labels = label_dollar(scale = 1e-6, suffix = " M"),
    limits = c(
      quantile(movies$Earnings, 0.01),
      quantile(movies$Earnings, 0.99)
    )
  ) +
  labs(
    title = "Earnings",
    x     = "Earnings [USD]",
    y     = "Density"
  ) +
  theme_minimal(base_size = 11)

gridExtra::grid.arrange(
  p1, p2,
  p3,
  layout_matrix = rbind(c(1, 2),
                        c(3, 3)),
  top = "Distributions of key financial variables"
)
```

Both Budget and Box Office follow approximately log-normal distributions:
roughly symmetric on the log scale, but strongly right-skewed on the
original scale. The Earnings density plot confirms that the majority of films
post a positive result, though the left tail of loss-making productions is
clearly visible.

## Correlation Matrix

Before turning to the individual research questions, it is useful to survey
the pairwise associations between all major variables simultaneously. We use
**Spearman rank correlation**, which is robust to outliers and makes no
assumption of linearity. `Earnings` is deliberately excluded from the matrix
— since it is defined as Box Office − Budget, its Spearman correlation with
those two variables is structurally inflated and does not carry independent
information.

```{r correlation-matrix, fig.height=6, fig.width=7}
cor_mat <- cor(
  movies[, c("IMDb score", "Budget", "Box Office",
             "Oscar and Golden Globes nominations",
             "Oscar and Golden Globes awards",
             "Release year",
             "Actors Box Office %",
             "Director Box Office %")],
  method = "spearman",
  use    = "pairwise.complete.obs"
)

corrplot(
  cor_mat,
  method      = "square",
  type        = "upper",
  tl.cex      = 0.85,
  tl.col      = "black",
  addCoef.col = "black",
  number.cex  = 0.70,
  col         = colorRampPalette(c(col_loss, "white", col_profit))(200),
  mar         = c(0, 0, 1, 0),
  title       = "Spearman rank correlations"
)
```

Four associations stand out. The strongest relationship in the matrix is
**Budget ↔ Box Office** (ρ ≈ 0.75): higher production budgets are strongly
associated with higher theatrical revenues, reflecting the role of marketing
spend and production scale in driving audience attendance. **IMDb score
correlates moderately with nominations and awards** (ρ ≈ 0.40–0.45),
indicating that critical industry recognition and audience appreciation tend
to co-occur. **Nominations and awards** are very strongly correlated with
each other, as expected — winning an award requires nomination first.
Finally, **Release year shows a weak negative correlation with IMDb score**
(ρ < 0), meaning older films tend to receive marginally higher ratings, a
pattern explored in detail in a later section.

<div class="note-box">
**A note on statistical significance at large sample sizes.** With n ≈
3,900 observations, even correlations as small as |ρ| ≈ 0.03 will produce
statistically significant p-values. Throughout this analysis, **the
magnitude of ρ matters far more than its p-value.** We interpret effect
sizes using Cohen's benchmarks: |ρ| < 0.10 = negligible, 0.10–0.30 =
small, 0.30–0.50 = medium, > 0.50 = large.
</div>

---

# Why Do Some Films Post a Negative Profit?

We begin by establishing the scale of the phenomenon:

```{r loss-count}
n_loss   <- sum(movies$Earnings < 0)
pct_loss <- round(mean(movies$Earnings < 0) * 100, 1)
cat("Films posting a loss:", n_loss,
    sprintf("(%.1f%% of the dataset)\n", pct_loss))
```

Approximately **`r pct_loss`%** of films in the dataset recorded a negative
theatrical result. This figure should be interpreted with care: as noted
above, `Earnings` captures box office revenues only. Films that appear
loss-making here may have recovered their budgets through home video,
streaming rights, television licensing, or merchandising.

### Top 10 films by profit and loss

Before any aggregated analysis, it is instructive to inspect the films at
both extremes of the earnings distribution:

```{r top10-chart, fig.height=5}
top_profit <- movies %>%
  slice_max(Earnings, n = 10) %>%
  mutate(label = paste0(Movie, " (", `Release year`, ")"), result = "Profit")

top_loss <- movies %>%
  slice_min(Earnings, n = 10) %>%
  mutate(label = paste0(Movie, " (", `Release year`, ")"), result = "Loss")

bind_rows(top_profit, top_loss) %>%
  ggplot(aes(x = reorder(label, Earnings), y = Earnings / 1e6, fill = result)) +
  geom_col() +
  geom_hline(yintercept = 0, colour = "grey30", linewidth = 0.5) +
  coord_flip() +
  scale_fill_manual(values = c("Profit" = col_profit, "Loss" = col_loss)) +
  scale_y_continuous(labels = label_dollar(suffix = " M")) +
  labs(
    title    = "Top 10 films — largest profit and largest loss",
    subtitle = "Earnings = Box Office − Budget (theatrical revenues only)",
    x = NULL, y = "Earnings [USD M]", fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

The asymmetry is striking: the largest profits comfortably exceed \$700M,
while the largest losses are roughly an order of magnitude smaller. This
reflects the general structure of the industry — catastrophic flops are
bounded by the size of their budgets, whereas a blockbuster can generate
revenues many times its production cost.

## Does Genre Influence Financial Loss?

Comparing raw film counts by genre is a necessary first step, but
inherently misleading — Action, Comedy and Drama simply produce more films
in absolute terms, which mechanically generates more absolute losses:

```{r genre-count, fig.height=5.5}
movies %>%
  mutate(result = ifelse(Earnings >= 0, "Profit", "Loss")) %>%
  count(Genre, result) %>%
  ggplot(aes(x = reorder(Genre, n), y = n, fill = result)) +
  geom_col(position = "stack") +
  coord_flip() +
  scale_fill_manual(values = c("Profit" = col_profit, "Loss" = col_loss)) +
  labs(title = "Number of films by genre — profit vs loss",
       x = "Genre", y = "Number of films", fill = "Financial result") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

The meaningful comparison is the **proportion** of loss-making films within
each genre, which controls for genre size:

```{r genre-proportion, fig.height=5.5}
genre_summary <- movies %>%
  group_by(Genre) %>%
  summarise(pct_loss = mean(Earnings < 0) * 100, n_films = n(), .groups = "drop")

ggplot(genre_summary,
       aes(x = reorder(Genre, pct_loss), y = pct_loss, fill = pct_loss)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(round(pct_loss, 1), "%  (n=", n_films, ")")),
            hjust = -0.05, size = 3.2) +
  coord_flip(clip = "off") +
  scale_fill_gradient(low = col_profit, high = col_loss) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.22))) +
  labs(
    title    = "Loss rate by genre",
    subtitle = "Percentage of films with negative earnings | n = total films in genre",
    x = "Genre", y = "Loss rate [%]"
  ) +
  theme_minimal(base_size = 12)
```

Drama and Biography show the highest loss rates among well-represented
genres, likely reflecting a combination of lower commercial appeal and niche
audience bases alongside sometimes substantial production costs. The Romance
genre registers a 0% loss rate, but with only 4 films this is a sampling
artefact rather than a meaningful finding.

### Kruskal-Wallis test and Dunn post-hoc comparisons

To formally test whether earnings distributions differ significantly across
genres, we apply the **Kruskal-Wallis test** — the non-parametric analogue
of one-way ANOVA, which ranks all observations and tests whether the rank
distributions are the same across groups. If the global test is significant,
**Dunn's post-hoc test** with Benjamini-Hochberg (BH) false discovery rate
correction identifies which specific genre pairs drive the result.

> **H₀:** The earnings distribution is identical across all genres.
>
> **H₁:** At least one genre has a different earnings distribution.

```{r genre-kruskal}
kw <- kruskal.test(Earnings ~ Genre, data = movies)
kw
```

<div class="result-box">
**Kruskal-Wallis result.** χ²(`r kw$parameter`) = `r round(kw$statistic, 2)`,
p `r ifelse(kw$p.value < 0.001, "< 0.001", paste("=", round(kw$p.value, 3)))`.
We **reject H₀**: earnings distributions differ significantly across genres.
At least one genre has a statistically distinct earnings profile from the
rest. The Dunn post-hoc analysis below identifies which specific pairs are
responsible.
</div>

```{r genre-dunn}
dunn_result <- dunn.test(
  movies$Earnings,
  movies$Genre,
  method = "bh",
  altp   = TRUE,
  list   = FALSE
)
```

<div class="result-box">
**Dunn post-hoc result.** With BH correction applied, the pairs showing the
most pronounced statistically significant differences involve Drama and
Biography on one side and high-revenue genres such as Action and Animation
on the other. Genre is a statistically significant driver of earnings,
though effect sizes are moderate — genre alone cannot reliably predict
whether any individual film will be profitable.
</div>

## Does Release Year Influence Financial Loss?

```{r year-count, fig.height=4.5}
movies %>%
  mutate(result = ifelse(Earnings >= 0, "Profit", "Loss")) %>%
  count(`Release year`, result) %>%
  ggplot(aes(x = `Release year`, y = n, fill = result)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c("Profit" = col_profit, "Loss" = col_loss)) +
  labs(title = "Number of films by release year — profit vs loss",
       x = "Release year", y = "Number of films", fill = "Financial result") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

Both profitable and loss-making films have grown in absolute number since
the late 1990s, primarily reflecting a rapid expansion of film production
volume rather than a deterioration of industry economics. As with genres,
raw counts conceal the more informative picture — proportions:

```{r year-proportion, fig.height=4.5}
movies %>%
  group_by(`Release year`) %>%
  summarise(pct_loss = mean(Earnings < 0) * 100, n = n(), .groups = "drop") %>%
  filter(n >= 5) %>%
  ggplot(aes(x = `Release year`, y = pct_loss)) +
  geom_col(fill = col_loss, alpha = 0.75) +
  geom_smooth(method = "loess", span = 0.75, se = TRUE,
              colour = "black", fill = "grey70", alpha = 0.25, linewidth = 0.8) +
  labs(
    title    = "Loss rate by release year",
    subtitle = "Years with ≥ 5 films only | LOESS span = 0.75 | grey band = 95% CI",
    x = "Release year", y = "Loss rate [%]"
  ) +
  theme_minimal(base_size = 12)
```

The LOESS smoother does not reveal a sustained directional trend. The loss
rate fluctuates irregularly year to year and shows no systematic increase
over time, suggesting that the overall proportion of loss-making productions
has remained roughly stable across the decades covered by this dataset.

## Does an Excessive Budget Cause a Loss?

We first compare the budget distributions of profitable and loss-making films
using summary statistics:

```{r budget-summary}
cat("--- Profitable films ---\n")
summary(movies$Budget[movies$Earnings >= 0])
cat("\n--- Loss-making films ---\n")
summary(movies$Budget[movies$Earnings < 0])
cat(sprintf(
  "\nMedian budget — Profitable: $%.1fM | Loss-making: $%.1fM\n",
  median(movies$Budget[movies$Earnings >= 0]) / 1e6,
  median(movies$Budget[movies$Earnings <  0]) / 1e6
))
```

```{r budget-boxplot}
movies %>%
  mutate(result = ifelse(Earnings >= 0, "Profit", "Loss")) %>%
  ggplot(aes(x = result, y = Budget, fill = result)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.2) +
  scale_fill_manual(values = c("Profit" = col_profit, "Loss" = col_loss)) +
  scale_y_log10(labels = label_dollar(scale = 1e-6, suffix = " M")) +
  labs(
    title    = "Budget distribution by financial result",
    subtitle = "Y-axis on log scale | horizontal bar = median",
    x = "Financial result", y = "Budget [USD]"
  ) +
  guides(fill = "none") +
  theme_minimal(base_size = 12)
```

The medians are close and the IQR ranges overlap substantially, suggesting
that budget size alone is not a reliable predictor of financial success.
However, the upper tail of the profitable group extends further — very large
budgets appear somewhat more often among profitable films, indicating that
the largest productions have a better chance of generating revenues that
justify their costs.

### Normality check and Wilcoxon rank-sum test

Before choosing a test, we verify the normality of the budget distribution
in each group using the **Anderson-Darling test**. R's built-in
`shapiro.test()` cannot be applied here because it imposes a hard upper
limit of n ≤ 5,000. Furthermore, at n ≈ 3,900 any normality test has
near-perfect statistical power — even trivially small departures from
normality will produce significant results. The normality test here serves
solely to formally confirm that non-parametric methods are appropriate.

```{r budget-normality}
cat("=== Loss-making films (Anderson-Darling) ===\n")
ad.test(movies$Budget[movies$Earnings < 0])

cat("\n=== Profitable films (Anderson-Darling) ===\n")
ad.test(movies$Budget[movies$Earnings >= 0])
```

<div class="result-box">
**Anderson-Darling result.** Both groups yield p < 0.001, confirming
non-normality in both subgroups. We proceed with the non-parametric
**Wilcoxon rank-sum test**.
</div>

> **H₀:** The budget distributions of profitable and loss-making films do not differ.
>
> **H₁:** Profitable films have higher budgets than loss-making films (one-sided test).

```{r budget-wilcoxon}
wt <- wilcox.test(
  movies$Budget[movies$Earnings >= 0],
  movies$Budget[movies$Earnings <  0],
  alternative = "greater"
)
wt
```

```{r budget-effect-size}
n1   <- sum(movies$Earnings >= 0)
n2   <- sum(movies$Earnings <  0)
r_rb <- 1 - (2 * wt$statistic) / (n1 * n2)
cat(sprintf("Rank-biserial correlation r = %.3f\n", r_rb))
cat("Benchmarks: |r| < 0.10 negligible | 0.10-0.30 small | 0.30-0.50 medium | > 0.50 large\n")
```

<div class="result-box">
**Wilcoxon rank-sum result.** W = `r formatC(wt$statistic, format = "f", digits = 0)`,
p `r ifelse(wt$p.value < 0.001, "< 0.001", paste("=", round(wt$p.value, 4)))`.
We **reject H₀**: profitable films have statistically significantly higher
budgets than loss-making films. The rank-biserial effect size r =
`r round(r_rb, 3)` is `r effect_label(r_rb)`, meaning the difference is
real but modest in practical terms — consistent with the substantial
distributional overlap seen in the boxplot above.

**Methodological note.** The Wilcoxon rank-sum test formally assesses
stochastic dominance (the probability that a randomly selected profitable
film has a higher budget than a randomly selected loss-making one).
Interpreting this as a difference in medians additionally requires the
assumption that both distributions share the same shape, differing only in
location. Given the similar IQR widths visible above, this assumption is
plausible but is not formally tested.
</div>

## Does IMDb Rating Influence Earnings?

```{r imdb-earnings-density}
movies %>%
  mutate(result = ifelse(Earnings >= 0, "Profit", "Loss")) %>%
  ggplot(aes(x = `IMDb score`, fill = result)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("Profit" = col_profit, "Loss" = col_loss)) +
  labs(
    title    = "IMDb score distribution by financial result",
    subtitle = "Overlapping distributions indicate IMDb score is not a reliable predictor of earnings",
    x = "IMDb score", y = "Density", fill = "Financial result"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

The density curves for profitable and loss-making films are nearly
indistinguishable. The profitable group is shifted marginally to the right,
but the overlap is so extensive that IMDb score cannot serve as a meaningful
predictor of financial outcome. Films with low ratings can still turn a
profit at the right budget level or with effective distribution, while
critically praised films can fail commercially if they lack mainstream
appeal. It is also worth noting that IMDb ratings are given by viewers
*after* watching the film — they are a co-occurring reflection of perceived
quality, not a causal input to financial success.

## Does a Director's Track Record Reduce Loss Risk?

Using the director performance quartile groups computed during data
preparation, we compare earnings distributions across the four groups:

```{r director-boxplot}
movies %>%
  filter(!is.na(director_group)) %>%
  ggplot(aes(x = director_group, y = Earnings / 1e6, fill = director_group)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.15) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey30") +
  scale_fill_manual(values = c(col_none, col_nom, col_gold, col_profit)) +
  scale_y_continuous(labels = label_dollar(suffix = " M")) +
  labs(
    title    = "Earnings by director's historical box office success rate",
    subtitle = "Quartiles of Director Box Office % | dashed line = break-even",
    x = "Director track record", y = "Earnings [USD M]"
  ) +
  guides(fill = "none") +
  theme_minimal(base_size = 12)
```

```{r director-table}
movies %>%
  filter(!is.na(director_group)) %>%
  group_by(director_group) %>%
  summarise(
    n        = n(),
    Q1       = round(quantile(Earnings, .25) / 1e6, 1),
    Median   = round(median(Earnings) / 1e6, 1),
    Mean     = round(mean(Earnings) / 1e6, 1),
    Q3       = round(quantile(Earnings, .75) / 1e6, 1),
    `Loss %` = round(mean(Earnings < 0) * 100, 1)
  )
```

A weak gradient is visible: directors in the highest-track-record quartile
(Q4) have both a higher median earnings and a lower loss rate than those in
the lowest quartile (Q1). However, the IQR remains very wide within every
group — a strong historical record reduces, but clearly does not eliminate,
the risk of a loss-making production.

### Top 10 directors by median earnings

The chart below shows the ten most consistently profitable directors in the
dataset, restricted to those with at least five films to ensure the median
is based on a sufficiently stable sample:

```{r top-directors, fig.height=5}
movies %>%
  group_by(Director) %>%
  summarise(n_films = n(), median_earnings = median(Earnings) / 1e6,
            .groups = "drop") %>%
  filter(n_films >= 5) %>%
  slice_max(median_earnings, n = 10) %>%
  ggplot(aes(x = reorder(Director, median_earnings), y = median_earnings,
             fill = median_earnings)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0("$", round(median_earnings), "M  (n=", n_films, ")")),
            hjust = -0.05, size = 3.2) +
  coord_flip(clip = "off") +
  scale_fill_gradient(low = col_profit, high = col_nom) +
  scale_y_continuous(labels = label_dollar(suffix = " M"),
                     expand = expansion(mult = c(0, 0.28))) +
  labs(
    title    = "Top 10 directors by median earnings",
    subtitle = "Directors with ≥ 5 films in the dataset",
    x = NULL, y = "Median earnings [USD M]"
  ) +
  theme_minimal(base_size = 12)
```

The five-film threshold is an accepted rule of thumb for median stability,
but is admittedly arbitrary. Rankings for directors close to the cut-off
should be treated with appropriate caution.

---

# What Influences IMDb Ratings?

## IMDb Score Distribution

Before examining potential predictors, we inspect the distribution of the
outcome variable itself:

```{r imdb-distribution, fig.height=4}
ggplot(movies, aes(x = `IMDb score`)) +
  geom_histogram(binwidth = 0.2, fill = col_nom, colour = "white", alpha = 0.9) +
  geom_vline(xintercept = median(movies$`IMDb score`),
             linetype = "dashed", colour = col_loss, linewidth = 0.8) +
  annotate("text",
           x     = median(movies$`IMDb score`) + 0.15,
           y     = Inf,
           label = paste0("median = ", round(median(movies$`IMDb score`), 2)),
           vjust = 1.5, hjust = 0, size = 3.5, colour = col_loss) +
  labs(title = "Distribution of IMDb scores", subtitle = "Dashed line = median",
       x = "IMDb score", y = "Number of films") +
  theme_minimal(base_size = 12)
```

```{r imdb-summary}
summary(movies$`IMDb score`)
```

The distribution is approximately symmetric, with a median of
`r round(median(movies$"IMDb score"), 2)` and a mode in the 6–7 range.
The absence of pronounced skewness means that the IMDb score itself does not
present distributional concerns, though we continue to use Spearman
correlation throughout for consistency with the skewed financial variables.

## Does a Higher Budget Lead to Better Ratings?

```{r budget-imdb}
ggplot(movies, aes(x = Budget, y = `IMDb score`)) +
  geom_point(alpha = 0.2, colour = col_nom) +
  geom_smooth(method = "loess", span = 0.75, se = TRUE,
              colour = col_loss, fill = col_loss, alpha = 0.15, linewidth = 1) +
  scale_x_log10(labels = label_dollar(scale = 1e-6, suffix = " M")) +
  labs(title = "IMDb score vs Budget",
       subtitle = "X-axis on log scale | LOESS span = 0.75 | band = 95% CI",
       x = "Budget [USD]", y = "IMDb score") +
  theme_minimal(base_size = 12)
```

The LOESS smoother is nearly flat across virtually the entire budget range,
indicating that production spend has little bearing on how audiences rate a
film. A slight upward curve is visible at very high budgets (above \$100M),
but the scatter around the curve is enormous. Both micro-budget productions
and \$200M blockbusters span the full range of IMDb scores.

### Spearman correlation test

> **H₀:** No monotonic association between Budget and IMDb score.
>
> **H₁:** A monotonic association exists.

```{r budget-imdb-test}
budget_cor <- cor.test(movies$Budget, movies$`IMDb score`, method = "spearman")
budget_cor
cat(sprintf("\nEffect size: |rho| = %.3f -> %s\n",
            abs(budget_cor$estimate), effect_label(budget_cor$estimate)))
```

<div class="result-box">
**Spearman correlation result (Budget ~ IMDb score).** ρ =
`r round(budget_cor$estimate, 3)`,
p `r ifelse(budget_cor$p.value < 0.001, "< 0.001", paste("=", round(budget_cor$p.value, 4)))`.
The association is statistically significant but `r effect_label(budget_cor$estimate)`
in practical terms. Budget explains only a negligible share of the variation
in IMDb scores — consistent with the near-flat LOESS curve observed above.
</div>

## Does Box Office Success Influence Ratings?

```{r boxoffice-imdb}
ggplot(movies, aes(x = `Box Office`, y = `IMDb score`)) +
  geom_point(alpha = 0.2, colour = col_nom) +
  geom_smooth(method = "loess", span = 0.75, se = TRUE,
              colour = col_loss, fill = col_loss, alpha = 0.15, linewidth = 1) +
  scale_x_log10(labels = label_dollar(scale = 1e-6, suffix = " M")) +
  labs(title = "IMDb score vs Box Office",
       subtitle = "X-axis on log scale | LOESS span = 0.75 | band = 95% CI",
       x = "Box office [USD]", y = "IMDb score") +
  theme_minimal(base_size = 12)
```

The pattern here is somewhat more pronounced than for budget. The LOESS
curve traces a modest positive trend — films that attracted larger audiences
tend to receive marginally higher IMDb scores. This could reflect a selection
effect (widely seen films accumulate more ratings, including from fans who
sought them out precisely because they expected to enjoy them) or simply
the fact that commercial and critical success are not independent.

### Spearman correlation test

> **H₀:** No monotonic association between Box Office and IMDb score.
>
> **H₁:** A monotonic association exists.

```{r boxoffice-imdb-test}
bo_cor <- cor.test(movies$`Box Office`, movies$`IMDb score`, method = "spearman")
bo_cor
cat(sprintf("\nEffect size: |rho| = %.3f -> %s\n",
            abs(bo_cor$estimate), effect_label(bo_cor$estimate)))
```

<div class="result-box">
**Spearman correlation result (Box Office ~ IMDb score).** ρ =
`r round(bo_cor$estimate, 3)`,
p `r ifelse(bo_cor$p.value < 0.001, "< 0.001", paste("=", round(bo_cor$p.value, 4)))`.
The association is statistically significant and slightly stronger than for
budget, but still `r effect_label(bo_cor$estimate)` by Cohen's benchmarks.
Box office revenue and IMDb score co-vary, but the relationship explains
only a small fraction of the variance in ratings.
</div>

## How Do Nominations Influence Ratings?

```{r nominations-distribution}
cat("Distribution of nomination counts:\n")
table(movies$`Oscar and Golden Globes nominations`)
```

Over 70% of films in the dataset received no nominations whatsoever. Films
with six or more nominations form a small but highly distinctive group,
likely representing the most acclaimed productions in the dataset. To
visualise how ratings relate to nomination count, we group films into four
bands:

```{r nominations-violin, fig.height=5}
movies %>%
  mutate(
    nom_group = cut(`Oscar and Golden Globes nominations`,
                    breaks = c(-1, 0, 2, 5, Inf),
                    labels = c("0", "1-2", "3-5", "6+"))
  ) %>%
  ggplot(aes(x = nom_group, y = `IMDb score`, fill = nom_group)) +
  geom_violin(alpha = 0.7, trim = FALSE) +
  geom_boxplot(width = 0.12, fill = "white", outlier.shape = NA) +
  geom_jitter(width = 0.05, alpha = 0.15, size = 0.6, colour = "grey40") +
  scale_fill_manual(values = c(col_none, col_nom, col_gold, col_profit)) +
  labs(
    title    = "IMDb score by number of nominations",
    subtitle = "Violin = full distribution | inner box = median & IQR | dots = individual films",
    x = "Number of nominations", y = "IMDb score"
  ) +
  guides(fill = "none") +
  theme_minimal(base_size = 12)
```

The violin shapes tell a clear story: as nomination count increases, the
entire distribution shifts upward and narrows considerably. Films with no
nominations show a broad distribution centred around 6–6.5 with a long left
tail. The 6+ group, by contrast, is tightly concentrated between 7 and 8.5
with very few films below 7. Greater industry recognition is strongly
associated with both higher and more consistent audience ratings.

## Does Winning an Award Have a Stronger Effect Than a Nomination Alone?

```{r awards-distribution}
cat("Distribution of award counts:\n")
table(movies$`Oscar and Golden Globes awards`)
```

Films with five or more awards constitute a fraction of a percent of the
dataset — exceptional productions by any measure.

```{r awards-violin, fig.height=5}
movies %>%
  assign_award_group() %>%
  ggplot(aes(x = award_group, y = `IMDb score`, fill = award_group)) +
  geom_violin(alpha = 0.7, trim = FALSE) +
  geom_boxplot(width = 0.12, fill = "white", outlier.shape = NA) +
  geom_jitter(width = 0.05, alpha = 0.15, size = 0.6, colour = "grey40") +
  scale_fill_manual(values = c(
    "No nominations"      = col_loss,
    "Nominated, no award" = col_none,
    "At least 1 award"    = col_gold
  )) +
  labs(
    title    = "IMDb score by award and nomination status",
    subtitle = "Violin + boxplot + jitter | orange = no nominations, teal = nominated, gold = awarded",
    x = NULL, y = "IMDb score"
  ) +
  guides(fill = "none") +
  theme_minimal(base_size = 12)
```

The three distributions are cleanly ordered and clearly separated. Films
with no nominations have the widest spread, with a pronounced left tail
representing the many productions that received poor audience scores.
Nominated-but-not-awarded films occupy a well-defined intermediate position,
with the distribution shifted upward and the lower tail substantially
shortened. Award-winning films form the tightest cluster, centred around
7.5–8 with very few films below 7. **Winning an award is associated with a
stronger and more consistent uplift in IMDb scores than receiving a
nomination alone.**

---

# Are Industry-Recognised Films Longer?

```{r runtime-awards}
movies %>%
  assign_award_group(labels = c("No nominations", "Nominated, no award", "Awarded")) %>%
  ggplot(aes(x = award_group, y = `Running time`, fill = award_group)) +
  geom_violin(alpha = 0.7, trim = FALSE) +
  geom_boxplot(width = 0.12, fill = "white", outlier.shape = NA) +
  geom_jitter(width = 0.05, alpha = 0.12, size = 0.6, colour = "grey40") +
  scale_fill_manual(values = c(
    "No nominations"      = col_loss,
    "Nominated, no award" = col_none,
    "Awarded"             = col_gold
  )) +
  labs(title = "Running time by industry recognition",
       subtitle = "Violin + boxplot + jitter | colours correspond to award status",
       x = NULL, y = "Running time [min]") +
  guides(fill = "none") +
  theme_minimal(base_size = 12)
```

A clear monotonic pattern emerges across award status groups. Films with no
nominations cluster tightly in the 85–120 minute range — the standard
commercial runtime. Nominated films spread more broadly, with a higher
median and a longer right tail. Awarded films are the longest on average,
with substantial right-skewness that reflects the inclusion of epic
productions exceeding 150 minutes. This pattern is consistent with the idea
that narratively ambitious, longer-form films receive more critical
attention — though it is equally plausible that prestigious films simply
have more resources and creative latitude to exceed standard runtime
conventions.

---

# Are Newer Films Rated Lower?

```{r year-imdb-scatter}
ggplot(movies, aes(x = `Release year`, y = `IMDb score`)) +
  geom_point(alpha = 0.15, colour = col_nom) +
  geom_smooth(method = "loess", span = 0.75, se = TRUE,
              colour = col_loss, fill = col_loss, alpha = 0.15, linewidth = 1) +
  labs(title = "IMDb score by release year",
       subtitle = "LOESS span = 0.75 | band = 95% CI",
       x = "Release year", y = "IMDb score") +
  theme_minimal(base_size = 12)
```

The LOESS curve traces a gradual downward trend from the 1960s through to
2016. To make this pattern more legible and to expose the sample sizes
underlying it, we aggregate to decade level:

```{r year-median-decade, fig.height=4.5}
movies %>%
  mutate(decade = (`Release year` %/% 10) * 10) %>%
  group_by(decade) %>%
  summarise(median_imdb = median(`IMDb score`), n = n(), .groups = "drop") %>%
  filter(n >= 20) %>%
  ggplot(aes(x = factor(decade), y = median_imdb, fill = median_imdb)) +
  geom_col(show.legend = FALSE, width = 0.7) +
  geom_text(aes(label = paste0(round(median_imdb, 2), "\n(n=", n, ")")),
            vjust = -0.3, size = 3.2, lineheight = 1.1) +
  scale_fill_gradient(low = col_loss, high = col_profit) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Median IMDb score by decade",
       subtitle = "Decades with >= 20 films | n shown per bar",
       x = "Decade", y = "Median IMDb score") +
  theme_minimal(base_size = 12)
```

The decline from the 1960s–70s to the 2000s–10s is visible but requires
careful interpretation. The n values shown in the bars are essential context:
pre-1970 decades contain very few films, meaning those estimates rest on
small and highly selective samples. Films old enough and notable enough to
appear in this dataset and to have attracted meaningful numbers of IMDb
ratings are, by definition, a curated elite of classic cinema. More recent
decades are represented by thousands of productions of varying quality.
**The observed rating decline most plausibly reflects survivorship bias and
a change in the composition of the rating audience, rather than a genuine
decline in film quality over time.**

### Spearman correlation test

We apply Spearman correlation because release year is ordinal in nature, the
IMDb score is bounded to 1–10, the relationship may be non-linear, and the
method is robust to outliers.

> **H₀:** No monotonic association between release year and IMDb score.
>
> **H₁:** A monotonic association exists.

```{r year-imdb-test}
year_cor <- cor.test(movies$`Release year`, movies$`IMDb score`,
                     method = "spearman")
year_cor
cat(sprintf("\nEffect size: |rho| = %.3f -> %s\n",
            abs(year_cor$estimate), effect_label(year_cor$estimate)))
```

<div class="result-box">
**Spearman correlation result (Release year ~ IMDb score).** ρ =
`r round(year_cor$estimate, 3)`,
p `r ifelse(year_cor$p.value < 0.001, "< 0.001", paste("=", round(year_cor$p.value, 4)))`.
We **reject H₀**: there is a statistically significant negative association
between release year and IMDb score. The effect size is
`r effect_label(year_cor$estimate)` by Cohen's benchmarks — release year
accounts for a minimal fraction of the variance in ratings. The statistical
signal is real, but its cause is almost certainly a compositional artefact
(survivorship bias and audience demographics) rather than a genuine temporal
trend in film quality.
</div>

---

# Summary

This exploratory data analysis covered `r nrow(movies)` feature films and
investigated three main areas: the determinants of financial outcome, the
factors associated with IMDb ratings, and the characteristics of
industry-recognised productions.

**Financial outcome** is difficult to predict from any single observable
feature. A Kruskal-Wallis test confirmed that earnings distributions differ
significantly across genres (χ²(`r kw$parameter`) = `r round(kw$statistic, 2)`,
p < 0.001), with Dunn post-hoc comparisons identifying Drama and Biography
as genres with systematically lower earnings relative to high-revenue genres
such as Action and Animation. Budget is statistically significantly higher
among profitable films (Wilcoxon rank-sum test, p < 0.001), but the
rank-biserial effect size is modest (r = `r round(r_rb, 3)`) and the
distributional overlap is large — budget alone is not a reliable predictor
of profit. The ROI variable (`Box Office / Budget`) complements raw earnings
by capturing relative financial performance independently of budget scale.
Release year showed no sustained directional effect on the loss rate.

**IMDb scores** are weakly but significantly associated with both budget
(ρ = `r round(budget_cor$estimate, 3)`, negligible effect) and box office
revenue (ρ = `r round(bo_cor$estimate, 3)`, small effect). Far stronger
associations emerge with industry recognition: films receiving multiple
nominations and, especially, award wins achieve markedly higher and more
tightly clustered ratings. The rating advantage of award-winning films over
non-nominated films is clearly visible even without formal testing. Newer
films carry marginally lower ratings on average
(ρ = `r round(year_cor$estimate, 3)`, small effect), but this is most
plausibly explained by survivorship bias and audience demographic shifts
rather than any decline in production quality.

**Industry-recognised films** are systematically longer, better-rated, and
more financially variable than films without nominations. The co-occurrence
of high audience ratings and industry recognition is consistent across
multiple analyses, though — as is appropriate for exploratory work — the
direction of any causal relationship between these outcomes cannot be
determined from the data alone.

Taken together, the findings reinforce a recurring theme in the economics of
cultural industries: film success is the product of many interacting factors,
and no single variable — budget, genre, director track record, or release
year — is a reliable standalone predictor of either financial performance or
audience reception.
